---
title: "Eradication Tools - Eradication progress"
author: "Dave Ramsey & Michael Scroggie"
date: "17/11/2020"
output:
  md_document: default
  html_document: default
  word_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(tibble)
```

## Eradication progress app user guide

This shiny app is primarily to be used for assessing progress towards eradication of a pest population.  As an  eradication program progresses, and additional data become available, the app will provide updates on the estimate of the residual size of the pest population remaining or the residual proportion of occupied sites. This help file provides some guidance about the use of the app including the types of monitoring data as well as the associated models that can be used to undertake an eradication progress.  Before using this app, the practitioner should ideally be familiar with the material in the accompanying document `Quantitative decision support for eradication - a primer`. In particular, the practitioner should have read `Spatial considerations: Extent, Management Zones and Sampling Units` and `Assessing Progress`. 

### Assessing progress

Consideration of the primary removal method is of great importance because the data collected from this method will be used to generate population estimates. For valid estimates to be obtained, the primary removal method must be able to remove individuals at a rate faster than they can be replaced. This can be easily visualised using a cumulative catch curve, which plots the cumulative catch (removals) of individuals against the catch per unit effort (CPUE) (Figure 1).	  The negative of the slope of this relationship is an estimate of the detection rate (also known as the catchability coefficient).  If a clear negative relationship is not evident or there is evidence of a flattening of the relationship (non-linearity), then that suggests that the population might not be reducing at a rate high enough to achieve eradication or that the detection rate of individuals is declining.  If this occurs, then alternative removal techniques need to be used.  One problem with relying on the cumulative catch versus CPUE plot is that the relationship may suggest that all individuals have been removed, when in fact, eradication has not been achieved.  This occurs because the relationship in Figure 1 assumes that detectability is constant for the duration of removal activities when in fact, the residual population is no longer susceptible to the removal method, (e.g. they may be trap-shy and therefore never get caught, resulting in detectability declining to zero). Identifying this situation depends on the use of a monitoring protocol running in parallel with the primary removal method that is able to detect residual survivors (e.g. cameras or scat-detecting dogs). 

```{r,fig.height=4,fig.width=4,fig.align='center',echo=FALSE,fig.cap="Figure 1. Plot of the relationship between CPUE and cumulative catch"}
catch<- c(40,13,9,8,3,0,0,2,1,1)
m<- coef(lm(catch/100 ~ cumsum(catch)))
plot(cumsum(catch), catch/100,  type="p", col="red",pch=16, xlab="Cumulative Catch",ylab="CPUE")
abline(a=m[1], b= m[2])

```

### Models

The models in the Eradication Progress tool are based primarily on removal data.  This means that the models require data on the total catch of individuals of the pest species removed, at each sampling location, for each removal phase (or "session").  Here a “session” refers to a distinct phase of time where removal of the pest is actively occurring.  For example, rodent trapping might be undertaken over 20 consecutive nights and then repeated again 6 months later.  Hence, there are two sessions with each session consisting of 20 nights trapping (i.e. 20 periods).  The distinction between “sessions” and “periods” within sessions is important as it underlies the data format required by the models used here.  Populations are assumed to be closed within a session.  That means, within a session, the population is not subject to births, (natural) deaths, immigration or emigration and the only change to the population is caused by the removal process over each period.  However, between sessions, populations can be assumed to be subject to these natural additions and/or losses.  The Eradication Progress tool contains models which cater for both these scenarios.

In general, the Eradication Progress tool relies on three different types of data.  1) spatial data of the region of interest, 2) spatial data for variables that might influence habitat preferences of the pest species 3) locations of removal devices (i.e. trap locations) and 4) records of the number of pest removals for each location, session and period.   The tool also caters for aspatial data which does not require device or pest locations.  This sort of data is usually collected during search-encounter type removals where there are no set device locations (e.g. aerial shooting from a helicopter).  All models used here require data from at least three (secondary) periods within a session.  Below we outline each of the data model types available within the Eradication Progress tool.  


* `Non-spatial removals`
	+ This model implements the catch-effort model of Gould & Pollock (1996). It is an aspatial model which means that removals are aggregated over all removal devices for each period.  This means that device locations and habitat information are not used or required for this model.  The catch and effort data are assembled from the removal histories by summing removals from each device, separately for each period. Effort data can be calculated as the number of devices set or the amount of search effort expended in each period. 
	
* `Non-spatial removals + index data`
	+ This model is the same as the non-spatial removal model detailed above but also allows index data to be included in the joint likelihood.  The index data can be any index of abundance (e.g. no. of detections on a camera etc.).  Index data can be useful for detecting the presence of individuals when the removal methods fails to detect any.  The idea is that the index should decrease proportionally as pest density decreases due to the removal method.  Index data and its associated effort (e.g. camera nights) should be included as extra columns labeled `index` and `ieffort`, respectively.   
	
* `Spatial removals (single session)`
	+ This model implements the multinomial removal model of Haines (2018).  Unlike the non-spatial model, device locations are required for this model and habitat information (`*.tif`) can be used to model spatial variation in initial abundance. Device locations are assumed to be independent. 
	
* `Spatial removals with index data (single session)`
	+ This model estimates the generalised removal model of Dorazio et al. (2005). However, it also has the facility to include additional index monitoring data into the analysis in addition to the removal data.  The index monitoring data can be derived from any index of abundance and are assumed to be collected in the same general vicinity (or a subset thereof) as the removal devices.  Hence, the index data should have the same number of rows and columns as the removal data and are uploaded into the app using `detection histories` button.  Removal devices without an associated monitoring device should have an `NA` inserted for the appropriate row in `detection histories`. 
	
* `Spatial removals (Multi session)`
	+ This model generalises the single session removal model of Haines (2018) to multiple sessions. This is achieved by 'stacking' data from multiple sessions and including an additional variable to estimate session-specific parameters or a single trend.  Detection probability is currently assumed to be the same for each session in the app but the model does allow session specific estimates of detection. This model is useful for estimating session specific abundance estimates without attempting to estimate how abundance changed (i.e. no explicit dynamic model parameters such as survival or recruitment).
	
* `Spatial presence/absence (Multi session)`
	+ This model estimates the dynamic multi-season occupancy model of MacKenzie et al. (2003) and only requires recording of  presence/absence data at each spatial location.  However, the model can also accept count data as input and will truncate this internally to presence/absence data.  This model generates estimates of the 'occupancy' probability for each site and will be most useful towards the final stages of the eradication program.  In the final stages when eradication appears likely, abundance becomes less important and focus shifts to the proportion of the region occupied.  The model is dynamic and generates estimates of the 'extinction' and 'colonisation' rate. However, at present the apps assume that these are constant and do not vary by session. Future versions of the app will allow explicit models for the colonisation and extinction parameters.     
	
### Inputs

All models in the progress app require inputs in particular formats.  For a more complete description of data formats, refer to the document `eradication_progress_data_req.pdf`:

* `catch and effort data`
	+ For the non-spatial models, the catch and effort data are assembled by summing removals for each period over the entire region of interest, separately for each session. Effort data is calculated similarly.  The required data should be entered into a spreadsheet (`*.csv` file) with columns `catch` recording the removals for each period, `effort` recording the trapping/search effort during the period and optionally `session` recording the session number (for multi-session data).  If data are for a single session, this column can be omitted.  If index data and associated effort are also available for each of the removal periods then it should be included as extra columns labeled `index` and `ieffort`, respectively.
	
* `Region boundary`
	+ Information is required about the region of interest where the eradication operation is being undertaken.  The minimal requirement is geo-spatial vector data of the region boundary either in ESRI shapefile format, packaged in a ZIP file or in the non-proprietary GeoPackage format (`*.gpkg`).  The boundary can comprise single or multiple polygons and can contain ‘holes’ (i.e. due to the presence of non-habitat such as a lake).  The region should be in a projected coordinate reference system (crs) with units in meters (m).
	
* `Habitat raster`
	+ Information on biophysical variables that might influence pest densities can be important information that can help delineate areas of potential high risk of infestation by the pest species.  This information can be supplied in the form of one or more raster files, one for each variable.  Habitat rasters can also comprise derived variables, such as outputs from species distribution models (SDMs).  Raster data should be supplied in GeoTiff format (`*.tif`) using the same coordinate reference system as the region shapefile above with units in meters.  The raster can be any resolution however care should be observed for large regions with a small resolution resulting in large files that may overwhelm available memory  
	
* `Trap locations`
	+ For the spatial models, a spreadsheet (`*.csv file`) recording the locations of each device is required. This dataset should have device locations coordinates with `X` (easting) and `Y` (northing) in the first and second columns, respectively.  The coordinate reference system should be the same as the shapefile of the region boundary.  This dataset should also have the same number of rows as the removal histories data.
	
* `Removal histories`
	+ The removal data should be in a spreadsheet (`*.csv` file) with removals for each period in columns and devices in rows.  Hence, an operation using 50 traps set for 6 nights would have corresponding removal data organized in 50 rows and 6 columns.
	
* `Detection histories`
	+ The index monitoring data are assumed to be derived from monitoring devices set in the same general vicinity (or a subset thereof) as the removal devices (e.g. camera traps.  Hence, the additional monitoring data should have the same number of rows and columns as the removal data. The index data can be any index of abundance derived from monitoring data (e.g. no. of detections on a camera etc.).  Removal devices without an associated index monitoring device should have an `NA` inserted in the appropriate row in `detection histories`.


## Analysis

Once the data for a particular model has been uploaded, the following functions can be performed:

* `plot map` 
	+ will produce a map of the region, plotted on background topographic or satellite info for the region.  Device locations and habitat raster data can also be plotted. 
	
* `plot removals` 
	+ will produce plots of the CPUE versus cumulative catch similar to figure 1. For multi-session data a separate plot is produced for each session.  
	
* `fit model` 
	+ will run the selected model and print the output in the "fitted model" pane. Outputs include the parameter estimates and a prediction of both the initial population size within the region based on the monitored sites as well as an estimate of the residual population size following the last removal period.  Any habitat covariates uploaded previously should also be available and can be selected in the model if desired.  
	
* `Estimate density surface` 
	+ will displays a map of the predicted relative density across the region. The relative density can be calculated for either the initial population (i.e. before removals) or the residual population (i.e. following the final removal occasion). This is most useful if habitat variables have been included in the model. 

### References

Dorazio, R. M., Jelks, H. L., & Jordan, F. (2005). Improving Removal-Based Estimates of Abundance by Sampling a Population of Spatially Distinct Subpopulations. Biometrics, 61(4), 1093–1101.

Gould, W. R., & Pollock, K. H. (1997). Catch-effort maximum likelihood estimation of important population parameters. Canadian Journal of Fisheries and Aquatic Sciences, 54, 890–897.

Haines, L. M. (2019). Multinomial N-mixture models for removal sampling. Biometrics, September, 1–9. https://doi.org/10.1111/biom.13147

MacKenzie, D. I., Nichols, J. D., Hines, J. E., Knutson, M. G., & Franklin, A. B. (2003). Estimating site occupancy, colonization, and local extinction when a species is detected imperfectly. Ecology, 84(8), 2200–2207. doi:10.1890/02-3090
